Columnar VCF using Zarr

### 2024-03-28 paper.tex

\subsection{Storing genetic variation data}
Storage is the foundation on which any large-scale data-driven computation
rests: how we encode and organise data on persistent storage (the format)
fundamentally affects and limits the efficiencies of downstream calculations.
Different formats are optimised for different purposes, and there are
often many competing requirements when designing a data storage format.
% TODO is there some literature we could cite on this? I'm making up the
% term "computational format" here, presumably there is lots  prior thought
% on these ideas?
For our  purposes here, it is important to distinguish between \emph{archival}
and \emph{computational} formats. An archival format is optimised for
maximum interoperability and future-proofing, and uses the simplest and
most robust technologies to ensure the long-term usability of the data.
Reasonable compression levels are important because
of storage and transfer costs, but the efficiency of data access is a
secondary concern.
A computational format, on the other hand, is optimised
for calculations, for \emph{using} the data, today. A computational format
is designed with particular data access patterns in mind, ultimately
to make a broad range of queries and downstream computations as efficient as
possible.

The SAV format is an extension of BCF that takes advantage
of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
They provide a command line interface and a C++ API.

@@

High level discussion of
strategies used to store genotype matrix and accompanying metadata.
Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\subsection{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks. We discuss parallelism, and why
bcftools and genozip can make limited use of available cores.
Also that single-machine parallelism is inevitably constrained
by IO (which is why savvy's thread utilisation drops off).
% Figure: \ref{fig-whole-matrix-compute}. Also
% supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.

@@

\subsection{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Discuss why storing AF values for crude population groupings is
intrinsically bad, and that it should be easy to do calculations
on aribitrary subsets. Also participants regularly drop-out of
large studies like UKB - do you update the AF values each time?

@@

\subsection{Non genotype data}

@@

@@

@@

% TEXT from HSG chapter section on compressing data

>>>>

\subsection{Storing the genotype matrix}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression ratios
and query/compute performance are well known, and many methods 
have been suggested to improve on these properties.[REVIEW METHODS]
High level discussion of
strategies used to store genotype matrix and accompanying metadata.

@@

Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\subsection{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks.

@@

\subsection{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Refer to fig~\ref{fig-subset-matrix-compute}.

@@

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of vcfzarr on a large human dataset.
[DETAILS ABOUT GeL DATA]
We converted the data to the intermediate columnar format 
which took X hours with a cluster of Y workers using the 
distributed explode commands. When then converted to Zarr using 
SOMETHING which took X hours using [RESOURCES]. In total, this gave 
X TiB of data, spread across Y thousand files, a compression of 
X over the original VCFs.
The detailed breakdown of per-column compression ratios is 
shown in Table~\ref{tab-genomics-england-data}.

@@

Table~\ref{tab-genomics-england-data} shows that the dataset size is 
dominated by a few columns. This is well known and has been remarked 
on several times [e.g. spVCF paper, etc]. [In this paragraph we discuss
how some of these fields can be considerably compressed with very
little loss of information. For example, the maximum Allele Depth (AD)
value seen in this dataset is >3000. However, when we compute the 
histogram of AD values we can see that something like 0.1\% of 
AD values are > 127, so we can reasonably store these as 8 bit 
ints. This reduces storage to X. Finish up by saying that more 
could be done, but it's beyond the scope of this paper. Maybe note 
that this kind of per-field analysis is easy now, e.g. computing 
the hist took X minutes using xarray and Dask on a cluster with
Y workers..]

To illustrate the difference in query performance we ran some 
\texttt{bcftools query} commands based on the Genomics England 
user guides. For example, finding all variants passing 
quality filters with an allele frequency X takes just 
x seconds on a single core (y seconds of which is writing the 
CSV file). The same query took X hours on one chunk, giving a 
an approximate total time of XX hours over all Y chunks 
in chromosomes 2.

@@

\section{Choosing default compressor settings}
To inform the choice of compression settings across different fields 
in VCF data, we analysed their effect on compression ratio on 
data from the 1000 Genomes project.

% [FILL IN WITH Shadi's analysis from here:
% https://github.com/sgkit-dev/bio2zarr/discussions/74
% Discussed in:
% https://github.com/sgkit-dev/vcf-zarr-publication/issues/26

@@

Storage is the foundation on which any large-scale data-driven computation
rests: how we encode and organise data on persistent storage (the format)
fundamentally affects and limits the efficiencies of downstream calculations.
Different formats are optimised for different purposes, and there are
often many competing requirements when designing a data storage format.
% TODO is there some literature we could cite on this? I'm making up the
% term "computational format" here, presumably there is lots  prior thought
% on these ideas?
For our  purposes here, it is important to distinguish between \emph{archival}
and \emph{computational} formats. An archival format is optimised for
maximum interoperability and future-proofing, and uses the simplest and
most robust technologies to ensure the long-term usability of the data.
Reasonable compression levels are important because
of storage and transfer costs, but the efficiency of data access is a
secondary concern.
A computational format, on the other hand, is optimised
for calculations, for \emph{using} the data, today. A computational format
is designed with particular data access patterns in mind, ultimately
to make a broad range of queries and downstream computations as efficient as
possible.

The SAV format is an extension of BCF that takes advantage
of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
They provide a command line interface and a C++ API.
% TEXT from HSG chapter section on compressing data
